Adaptive evolution with aneuploidy and mutation

### 2022-02-21 ms/ms.tex
clean up todos
mended some TODOs
update MP todos
results regarding only the aneuploidy trajectory

\paragraph*{The present study.}
Here, we develop evolutionary-genetic models that include the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
These models follow a population of cells characterised by both their ploidy and their genotype.
We fit these models to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to perform model selection between different models, thereby testing hypotheses about the evolutionary process. 
We analyze these evolutionary-genetic models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
We find that \textbf{TODO} % TODO

@@

\paragraph{Model comparison.} We perform model selection using WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta) &\ =\ 
-2\log\mathbb{E}[\likelihood(\theta)]\ +\ 2\mathbb{V}[\log\likelihood(\theta)]
\end{aligned}
\end{equation}
where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution.
WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
% TODO Describe posterior plots comparison method - Fig 4 & S3

@@

Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
We tried to use additional growth curves. We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$. The same prior was used for $w_{2n*}/w_{2n+1*}$. This prior resulted in WAIC 0.32, compared to 0.27 with the above informative prior.
% TODO what are the results ?
We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\cref{fig:growth-curves}-B) to estimate $1-c$ . This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
% TODO it make sense because bla bla (to explain)
Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In both cases the algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\cref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
mutation rate, $\mu=2.93\substack{+1.122 \\ -2.824}\times10^{-6}$,
aneuploidy rate, $\delta=1.716\substack{+0.974 \\ -0.384}\times10^{-3}$,
and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.001}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
% TODO we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
This estimated aneuploidy rate agrees with previous estimates. % TODO refs
The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
%We can see from the posterior distribution~(\cref{fig:posterior}) the correlation between fitness of different genotypes. 
% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?

@@

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \cref{fig:ppc-plot}), which agrees with experimental results.
% TODO replace "roughly 300" with the exact number.
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\cref{fig:sensitivity}).

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). The estimated parameter values without aneuploidy are: 
%% generated with MAP-and-hdi.ipynb
mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
fitness of mutant, $w_{2n^*}=1.013\substack{+0.000 \\ -0.000}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\cref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO WAIC?

@@

Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) in the single-locus model and how it is affected by the model parameters. As a stochastic model, there are two events that can occur in the \euwt\ population. The first is a mutation that produces a \eumt\ cell. The second is a missegregation event that produces a \anwt\ cell. The probability with which these events occur is significantly different. We are interested in the average waiting time until one of the two events occurs and then lead to adaptation (in the second case, it will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt).
We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
\begin{subequations}
\begin{align}
p_{2n^*} &= 1-(1-2\mu \Delta w_{2n^*})^N \;, \\
p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \;,
\end{align}
\end{subequations}
where $\Delta w_i$ denotes the relative difference between the fitness of genotype $i$ and the population mean fitness.
However, we have two competing events: if one event occurs (e.g., mutation) then the other (aneuploidy) is unlikely to occur,  we need to find the minimum of these waiting times. The minimum of two geometric random variables with parameters $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with parameter $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
The probability that aneuploidy will be the first event to occur is
\begin{equation}
p_A = \frac{p_{2n+1}}{p_{2n^*}+p_{2n+1}} \;.
\end{equation}
We are interested in the expected time to fixation of \eumt, and the events leading to this fixation depend on the event that occurs first. If \eumt\ appears first and avoids extinction by drift, then it will fix. If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt. Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.
In general, we assume that the waiting time until a new genotype occurs is much longer than the time to its fixation. Nonetheless, these fixation times must be included in a rigorous analysis. There are two ways to compute the fixation times and they differ in the assumption of the underlying process.

A deterministic-process approach is given by the classical bi-allelic single-locus equation, which provides an approximation for the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$ is % TODO MP is this correct?
 % TODO MP ref for equation? Crow and Kimura?
\begin{equation}
\tilde T_{i \to j} = 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
\end{equation}
A stochastic-process approach for approximating $\tilde T_i$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}. 
Thus, the expected adaptation time via the aneuploidy trajectory is given by
\begin{equation}\label{eq:T_aneu}
 T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + \tilde T_{2n \to 2n+1} + \tilde T_{2n+1 \to 2n+1^*} + \tilde T_{2n+1^* \to 2n^*} \;,
 \end{equation}
the expected adaptation time via the mutation trajectory is given by 
\begin{equation}\label{eq:T_mut}
T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \tilde T_{2n \to 2n^*} \;,
\end{equation}
and the expected adaptation time is given by
 \begin{equation}\label{eq:T_total}
 T = p_A T_{A}+(1-p_A) T_{M}.
 \end{equation}

@@

To validate the approximation in \cref{eq:T_total}, we performed simulations of the full model (\cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\cref{fig:only mut}), and with the other parameters as inferred above. The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells. In that case, the expected number of new mutants is ~0.5, and therefore we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which is in contrast to our assumption that an adaptation event starts from a single mutant cell.

 \begin{figure}[h!]
 \centering
 \includegraphics[width=.49\textwidth]{../figures/figinvpNonlymut0.pdf}
 \includegraphics[width=.49\textwidth]{../figures/figinvpmuonlymut0.pdf}
 \caption{\textbf{Expected adaptation time without aneuploidy.}
  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B). Mean and 95\% CI from simulations in blue. Analytic approximation (\cref{eq:T_total}) in orange. Here, XXXX.}
  %TODO MP add panel labels A and B
  %replace XXXX with parameter values and remove them from the figure titles.
  %y label "Expected adaptation time"
  % remove legend
  % make it a single figure with two panels, then there is no need for y label on panel B
  % The error bars a quite large, can you make them smaller by running more simulations?
  % Add more x ticks for panel A (10^6, 5 10 ^6 ...)
  % maybe start x axis in panel A with lower population size? 10^4 or 10^5 ? does the approximation work well there?
 \label{fig:only mut}
 \end{figure}

@@

% Fig Curveball 
\begin{figure}[h]
	\centering
	% generated with growth_curves.ipynb and deg = '39' 
	\includegraphics[width=.95\textwidth]{../figures/evo39_fitness_39deg.pdf} 
	% generated with growth_curves.ipynb and deg = '30' 
	\includegraphics[width=0.95\textwidth]{../figures/evo39_fitness_30deg.pdf} 
	% generated with growth_curves_refined.ipynb and deg = '30' 
	\includegraphics[width=0.95\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
\caption{
    \textbf{Fitness estimation from growth curves.}
    % TODO caption should explain what each panel shows.
    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}: $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
    \textbf{(E-H)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{30}{\celsius}: $\hat w_{2n+1}/w_{2n}$= 0.986 (95\% CI: 0.979 - 0.994).
    \textbf{(I-L)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).\\
    Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
	Fitness estimated from growth curves using \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. See \emph{Models and Methods, Prior distributions} for more details.
} 
\label{fig:growth-curves}
\end{figure}

% Fig likelihood profile
%% generated by sensitivity-analysis.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-A.pdf}      
      \label{fig:sensitivity-mutation}
  \end{subfigure}
  \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-B.pdf}      
      \label{fig:sensitivity-aneuploidy}
  \end{subfigure}
  \\
   \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-C.pdf}      
      \label{fig:sensitivity-anwt}
  \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-D.pdf}      
      \label{fig:sensitivity-anmt}
  \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-E.pdf}      
      \label{fig:sensitivity-eumt}
  \end{subfigure}
  \caption{
    \textbf{Likelihood profiles.} Sensitivity of the approximate likelihood of the single-locus model, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates. Dashed vertical line represents the MAP value.
  }

  \label{fig:sensitivity}
  % TODO what defines the x range? is it the prior distribution range?
  %ANS: yes for mutation and aneuploidy rates. For fitness just around the MAP in range that looks beautiful.
\end{figure}

>>>>

\paragraph*{The present study.}
Here, we develop evolutionary-genetic models that include the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
These models follow a population of cells characterised by both their ploidy and their genotype.
We fit these models to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to perform model selection between different models, thereby testing hypotheses about the evolutionary process. 
We analyze these evolutionary-genetic models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
We find that \textbf{} % TODO

@@

\paragraph{Model comparison.} We perform model selection using WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta) &\ =\ 
-2\log\mathbb{E}[\likelihood(\theta)]\ +\ 2\mathbb{V}[\log\likelihood(\theta)]
\end{aligned}
\end{equation}
where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution.
WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
% TODO IK Describe posterior plots comparison method - Fig 4 & S3

@@

Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
We tried to use additional growth curves. We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$. The same prior was used for $w_{2n*}/w_{2n+1*}$. This prior resulted in WAIC 0.32, compared to 0.27 with the above informative prior.
We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\cref{fig:growth-curves}-B) to estimate $1-c$ . This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
% TODO IK it makes sense because bla bla (to explain)
Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In both cases the algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\cref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
mutation rate, $\mu=2.93\substack{+1.122 \\ -2.824}\times10^{-6}$,
aneuploidy rate, $\delta=1.716\substack{+0.974 \\ -0.384}\times10^{-3}$,
and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.001}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
% TODO IK we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
This estimated aneuploidy rate agrees with previous estimates. % TODO refs
The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
%We can see from the posterior distribution~(\cref{fig:posterior}) the correlation between fitness of different genotypes. 
% TODO IK say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?

@@

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \cref{fig:ppc-plot}), which agrees with experimental results.
% TODO IK replace "roughly 300" with the exact number.
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\cref{fig:sensitivity}).

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). The estimated parameter values without aneuploidy are: 
%% generated with MAP-and-hdi.ipynb
mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
fitness of mutant, $w_{2n^*}=1.013\substack{+0.000 \\ -0.000}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\cref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO IK WAIC?

@@

Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) in the single-locus model and how it is affected by the model parameters. As a stochastic model, there are two events that can occur in the \euwt\ population. The first is a mutation that produces a \eumt\ cell. The second is a missegregation event that produces a \anwt\ cell. The probability with which these events occur is significantly different. We are interested in the average waiting time until one of the two events occurs and then lead to adaptation (in the second case, it will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt).
We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt\ are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
\begin{subequations}
\begin{align}
p_{2n^*} &= 1-(1-2\mu \Delta w_{2n^*})^N \;, \\
p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \;,
\end{align}
\end{subequations}
where $\Delta w_i$ denotes the relative difference between the fitness of genotype $i$ and the population mean fitness.
However, we have two competing events: if one event occurs (e.g., mutation), then the other (aneuploidy) is unlikely to occur,  we need to find the minimum of these waiting times. The minimum of two geometric random variables with parameters $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with parameter $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
The probability that aneuploidy will be the first event to occur is
\begin{equation}
p_A = \frac{p_{2n+1}}{p_{2n^*}+p_{2n+1}} \;.
\end{equation}
We are interested in the expected time to fixation of \eumt, and the trajectory leading to this fixation depend on the event that occurs first. If \eumt\ appears first and avoids extinction by drift, then it will fix. If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt. Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.
In general, we assume that the waiting time until a new genotype occurs is much longer than the time to its fixation. Nonetheless, these fixation times must be included in a rigorous analysis. There are two ways to compute the fixation times and they differ in the assumption of the underlying process.

A deterministic-process approach is given by the classical bi-allelic single-locus equation, which provides an approximation for the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$ is % TODO MP is this correct?
 % TODO MP ref for equation? Crow and Kimura?
\begin{equation} \label{eq:T_fix_x}
\tilde T_{i \to j} = 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
\end{equation}
A stochastic-process approach for approximating $\tilde T_i$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}. 
Thus, the expected adaptation time via the aneuploidy trajectory is given by
\begin{equation}\label{eq:T_aneu}
 T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + \tilde T_{2n \to 2n+1} + \tilde T_{2n+1 \to 2n+1^*} + \tilde T_{2n+1^* \to 2n^*} \;,
 \end{equation}
the expected adaptation time via the mutation trajectory is given by 
\begin{equation}\label{eq:T_mut}
T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \tilde T_{2n \to 2n^*} \;,
\end{equation}
and the expected adaptation time is given by
 \begin{equation}\label{eq:T_total}
 T = p_A T_{A}+(1-p_A) T_{M}.
 \end{equation}

@@

As a first step, to validate the approximation in \cref{eq:T_total}, we performed simulations of the full model (\cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\cref{fig:only mut}), and with the other parameters as inferred above. The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells. In that case, the expected number of new mutants is ~0.5, and therefore we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which is in contrast to our assumption that an adaptation event starts from a single mutant cell.

 \begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{../figures/figgrid_noaneu.pdf}
 \caption{\textbf{Expected adaptation time without aneuploidy.}
  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B). Mean and 95\% CI from simulations in blue. Analytic approximation (\cref{eq:T_total}) in orange. Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$. In panel A, $\mu=9.6\cdot 10^{-8}$ and in panel B, $N=6\cdot 10^6$.
  }
  %TODO MP add panel labels A and B
  %replace XXXX with parameter values and remove them from the figure titles.
  %y label "Expected adaptation time"
  % remove legend
  % make it a single figure with two panels, then there is no need for y label on panel B
  % The error bars a quite large, can you make them smaller by running more simulations?
  % Add more x ticks for panel A (10^6, 5 10 ^6 ...)
  % maybe start x axis in panel A with lower population size? 10^4 or 10^5 ? does the approximation work well there?
 \label{fig:only mut}
 \end{figure}

 \begin{figure}[h!]
 %\centering
 \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr0.jpg}
 \caption{\textbf{Expected adaptation time without mutations from $\euwt$ to $\eumt$.}
  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C). Mean and 95\% CI from simulations in blue. Analytic approximation (\cref{eq:T_total}) in green. The red curve is achieved by setting $\tilde{T}_{\euwt \rightarrow \anwt}=\tilde{T}_{\anwt \rightarrow \anmt}=0$. The orange line is achieved by setting $x=0.017$ in \cref{eq:T_fix_x} for the fixation time of $\eumt$, such that it is well fitting in panel B. Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$. Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.  }
 \label{fig:no direct mut}
 \end{figure}

 Second, we ran simulations without state changes from $\euwt$ to $\eumt$. Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. Here, the aneuploid states are transient and fixation could be outpaced by the arrival of the next type. This would explain the discrepancy between the green curves in \cref{fig:no direct mut}. To asses this further, \cref{fig:fixT vs waitT} provides a clear picture. At least for intermediate to high values of the respective range of the parameters, the waiting time until a succesful new type appears is far smaller than that to fixation of the old type. Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in the better fitting red curves in \cref{fig:no direct mut}. Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C. However, we stop simulations in state $\eumt$ if they have frequency $0.95$.

  The discrepancies in \cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant and the simulations decreas log-linearly with increasing $\delta$.

 \begin{figure}[h!]
 %\centering
 \includegraphics[width=1\textwidth]{../figures/figgrid_fixT_vs_waitT.jpg}
 \caption{\textbf{Expected time of fixation and waiting time events on the aneuploid trajectory.}
  The expected time to appearance and fixation of each step seperately in the aneuploid trajectory as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C). Solid curves give the fixation time and dashed curves the waiting time. Curves of the same color indicate the same base population state;  e.g. the blue dashed curve is the waiting time in state $\anwt$ until a succesful individual of type $\anmt$ appears. The corresponding solid curve indicates the time to fixation of the population in state $\anwt$. Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$. Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.  }
 \label{fig:fixT vs waitT}
 \end{figure}

In summary, whereas the two possible trajectories to $\eumt$ do not compete against each other in both \cref{fig:only mut} and \cref{fig:no direct mut}, the approximations are doing well only in restricted parameter ranges. These tend to be, where the respective parameters are rather small. If they grow bigger, they are violating some assumptions of our approximations.

@@

% Fig Curveball 
\begin{figure}[h]
	\centering
	% generated with growth_curves.ipynb and deg = '39' 
	\includegraphics[width=.95\textwidth]{../figures/evo39_fitness_39deg.pdf} 
	% generated with growth_curves.ipynb and deg = '30' 
	\includegraphics[width=0.95\textwidth]{../figures/evo39_fitness_30deg.pdf} 
	% generated with growth_curves_refined.ipynb and deg = '30' 
	\includegraphics[width=0.95\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
\caption{
    \textbf{Fitness estimation from growth curves.}
    % TODO IK caption should explain what each panel shows.
    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}: $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
    \textbf{(E-H)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{30}{\celsius}: $\hat w_{2n+1}/w_{2n}$= 0.986 (95\% CI: 0.979 - 0.994).
    \textbf{(I-L)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).\\
    Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
	Fitness estimated from growth curves using \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. See \emph{Models and Methods, Prior distributions} for more details.
} 
\label{fig:growth-curves}
\end{figure}

% Fig likelihood profile
%% generated by sensitivity-analysis.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-A.pdf}      
      \label{fig:sensitivity-mutation}
  \end{subfigure}
  \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-B.pdf}      
      \label{fig:sensitivity-aneuploidy}
  \end{subfigure}
  \\
   \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-C.pdf}      
      \label{fig:sensitivity-anwt}
  \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-D.pdf}      
      \label{fig:sensitivity-anmt}
  \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{../figures/sensitivity-E.pdf}      
      \label{fig:sensitivity-eumt}
  \end{subfigure}
  \caption{
    \textbf{Likelihood profiles.} Sensitivity of the approximate likelihood of the single-locus model, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates. Dashed vertical line represents the MAP value. The prior distributions for the mutation rate and aneuploidy rate are $\mu \sim U(10^{-9}, 10^{-5})$ and $\delta \sim U(10^{-6}, 10^{-2})$, respectively. 
  }

  \label{fig:sensitivity}
\end{figure}
